---
title: "DESeq on Heat Shock"
author: "Tolga Tabanlı"
date: "2024-12-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=F, message=F}
library(DESeq2)
library(tidyverse)
library(edgeR)
library(ggpubr)
library(plotly)
```

# Reading the hat shock data matrix and the mapping for the samples.

```{r, warning=F, message=F}
colData <- read_tsv("sample_mapping.tsv", col_names = T) %>%
  mutate(across(everything(), as.factor)) %>%
  column_to_rownames(var = "sample")

heatshock_counts <- read_tsv(here::here("complex_yeast_heatshock.tsv"), col_names = T) %>%
  column_to_rownames(var = "gene_id") %>%
  relocate(rownames(colData))

knockouts <- colData %>%
  select(knockout) %>%
  distinct() %>%
  pull() %>%
  relevel("Wildtype")
```

Following is the function used to apply deseq() on with the given parameters
and saving the plots. All are compared with 25C and t = 0 as reference with
their respective strains/knockouts.

```{r, warning=F, message=F}
# returns two plots comparing with temp=25 and t=0 of the respective strain
multiple_deseq <- function(countData, colData, knock, temp) {
  subColData <- colData %>%
    filter(knockout == knock) %>%
    filter((temperature == 25 & time == 0) | temperature == temp)
  
  plots <- list()
  i = 1
  for (t in c(10, 30)) {
    subsample <- subColData %>%
      filter(time %in% c(0, t))
    
    subcounts <- heatshock_counts %>% select(rownames(subsample))
    
    dds <- DESeqDataSetFromMatrix(countData = subcounts,
                                  colData = subsample,
                                  design = ~ time)
    deseq <- DESeq(dds, quiet = TRUE)
    res <- results(deseq)
    res <- res %>%
      pluck("listData") %>%
      as_tibble() %>%
      drop_na()
    plots[[i]] <- ggplot(data = res,
                       aes(x = log2FoldChange, y = -log10(pvalue),
                           colour = padj < 0.05)) +
      geom_point() +
      # scale_x_continuous(limits = c(-3,3)) +
      # scale_y_continuous(limits = c(0, 20))
      labs(title = paste(knock, temp, "C, t =", t))
    i <- i + 1
  }
  return(plots)
}
```

# DESeq

Here, the plots from the DESeq are drawn according to temperature and time.
The titles of the plots indicate the group compared against wildtype together
with the temperature and time info.

```{r, warning=F, message=F, fig.width = 10, fig.height = 30}
# DESeq
temps <- c(37, 42)
knockouts <- colData %>%
  select(knockout) %>%
  unique() %>%
  pull()

combs <- expand.grid(temperature = temps, knockout = knockouts)
plots <- list()
for (row in 1:nrow(combs)) {
  plots[[row]] <- multiple_deseq(heatshock_counts,
                                 colData,
                                 combs[row,"knockout"],
                                 combs[row,"temperature"])
}

plots <- unlist(plots, recursive = FALSE)
ggarrange(plotlist = plots, ncol = 2, nrow = 8, common.legend = T)

```

