---
title: "AlphaFold IDPs"
author: "Tolga Tabanli"
date: "2025-01-25"
output: pdf_document
knit: (function(input, encoding) {
  rmarkdown::render(input,
                    encoding = "UTF-8",
                    output_dir = "/home/tolga/Desktop/IDPs_under_Heatshock/Knittings")})
---

```{r setup-aiupred, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/home/tolga/Desktop/IDPs_under_Heatshock/')
library(tidyverse)
```

## AIUPred

The disorder scores were calculated using the AIUPred's Python package.
Basic plots to explore the score distributions were hence also done
by matplotlib's pyplot. So instead, I will continue with decision/filtering.

### Accession ID Conversion

The database used for input to AIUPred was UniProt. However, the counts
data for the DESeq analysis uses proteins with Ensembl IDs.
This is why I used biomaRt library to do this conversion.

```{r prep-aiupred, eval=F, message=F}
library(tidyverse)
library(biomaRt)

# for recreating biomart
ensembl_up_mapping <- useEnsembl(biomart = "genes",
                                 dataset = "scerevisiae_gene_ensembl") %>%
  getBM(attributes = c("ensembl_gene_id", "uniprotswissprot"),
        mart = .)

# use prepared mapping from biomart
ensembl_up_mapping <- readRDS("IDP decisions/ensembl_up_mapping.Rds")
```

Finally, the disorder modes of the proteins were read from the TSV file
that was generated through python.
Proteins were then filtered around the cut-off of 0.6.

```{r, eval=F}
aiupred_all_proteins_mode <- read_tsv(here::here("aiupred/aiupred_modes_of_proteins.tsv"))

aiupred_all_proteins_mean %>%
  inner_join(ensembl_up_mapping, join_by(uniprot == uniprotswissprot)) %>%
  dplyr::select(ensembl_gene_id, mean) %>%
  dplyr::rename("mean_disorder" = "mean") %>%
  write_tsv("IDP decisions/all_proteins_from_aiupred_with_mean.tsv")
```

