---
title: "AlphaFold IDPs"
author: "Tolga Tabanli"
date: "2025-01-25"
output: pdf_document
knit: (function(input, encoding) {
  rmarkdown::render(input,
                    encoding = "UTF-8",
                    output_dir = "/home/tolga/Desktop/IDPs_under_Heatshock/Knittings")})
---

```{r setup_alphafold, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/home/tolga/Desktop/IDPs_under_Heatshock/')
library(tidyverse)
```

## AlphaFold

In AlphaFold's .cif files, the local pLDDTs are given for each residue.
These were extracted using R's extensive regex capabilities.
To optimize the process, certain flags/patterns were used to exactly
locate the required segment.

```{r prep-alphafold, eval=F}
dir <- here::here("alphafold cifs")
files <- list.files(dir)

local_plddts <- tibble(uniprot = character(), scores = list())
start <- "_ma_qa_metric_local.ordinal_id"
end <- "#"
score_pattern <- "\\b\\d+\\.\\d+\\b"  # regex for floating point numbers
```

The file is first scanned until a start flag is encountered (found_start)
and then the pattern is searched until an end flag (end).

```{r, eval=F}
i <- 1
for (file in files) {
  print(i) # for progress

  accession <- str_split_1(file, "-")[2] # get accession from the file name
  file_connection <- file(paste0(dir, "/", file), open = "r")
  found_start <- FALSE
  scores <- c()
  while(TRUE) {
    line <- readLines(file_connection, n = 1)
    if(!found_start) {
      if (line == start) found_start <- TRUE
    } else {
      if (line == end) {
        break
      }
      # extract the decimal number and concatenate to the scores
      scores <- c(scores, as.numeric(regmatches(line, regexpr(score_pattern, line))))
    }
  }
  local_plddts <- local_plddts %>% add_row(uniprot = accession, scores = list(scores))
  i = i + 1
}
```


Next, the object was stored in an Rds for later access.

```{r, eval=F}
saveRDS(local_plddts, "IDP decisions/local_plddts.Rds") # Store

local_plddts <- readRDS("IDP decisions/local_plddts.Rds") # Load
```

```{r hidden-plddt-load, include=F}
local_plddts <- readRDS("IDP decisions/local_plddts.Rds")
```

### Accession ID Conversion

Next, the accessions of the proteins are converted to
Ensembl Gene IDs for compatibility with count data by using
a mapping that has been created by "biomaRt".

```{r, eval=F}
ensembl_up_mapping <- readRDS("IDP decisions/ensembl_up_mapping.Rds") # through biomaRt

local_plddts %>%
  group_by(uniprot) %>%
  mutate(mean_plddt = mean(unlist(scores))) %>%
  ungroup() %>%
  inner_join(ensembl_up_mapping, join_by(uniprot == uniprotswissprot)) %>%
  dplyr::select(ensembl_gene_id, mean_plddt) %>%
  write_tsv("IDP decisions/all_proteins_from_alphafold_with_mean.tsv")
```

### Distributions

For distribution of all local pLDDTs, the following code was used.
The data frame of proteins residue scores are first reduced to a
"pool" of all scores, which is then stored in csv file.
By reading this csv file, the data is also automatically enframed, which
can be given to ggplot as data. Saving steps are commented out
to avoid side effects.
The extensive design parameters had to be used for obtaining similarity to
Python's Matplotlib design.

```{r all-plddts-distro, fig.width=5, fig.align='center', eval=F}
all_scores <- local_plddts %>% select(scores) %>% pull() %>% reduce(c)
# write.csv(all_scores, "alphafold/local_plddts.csv")
all_scores <- read.csv("alphafold/local_plddts.csv") %>%
  `colnames<-`(c("row_names", "locals"))
# saveRDS(all_scores, "IDP decisions/all_local_plddt_scores.Rds")

# Hist of all local plddt scores
ggplot(all_scores, aes(locals)) +
  geom_histogram(bins = 100, fill = "#1f77b4") +
  xlab("Local pLDDT") +
  ylab("") +
  labs(title = "Distribution of all local pLDDTs from AlphaFold") +
  theme(axis.line = element_line(colour = "black"),
        axis.text = element_text(size=14),
        axis.title = element_text(size=14),
        title = element_text(size=12),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        plot.margin = margin(t = 20),
        aspect.ratio = 0.75) +
  scale_x_continuous(breaks = c(0,20,40,60,80,100)) +
  scale_y_continuous(expand = c(0,0,0.1,0)) +
  theme(plot.title = element_text(hjust = 0.5))
```

It was a harder challenge to find the modes of the proteins.
I first needed to define a function to create "intrinsic histograms" for each
protein. I had used a similar method in python with scipy's stats. (Hence the name)

```{r find-mode-plddt, eval=F}
find_mode_python <- function(scores, bins) {
  bin_edges <- seq(min(scores), max(scores), length.out = bins + 1)
  bin_counts <- hist(scores, breaks = bin_edges, plot = FALSE)$counts
  bin_centers <- (bin_edges[-1] + bin_edges[-length(bin_edges)]) / 2
  mode_value <- bin_centers[which.max(bin_counts)]
  return(mode_value)
}

# Calculate and save the modes
with_modes <- local_plddts %>% mutate(mode_plddt =
                          sapply(scores, find_mode_python, bins = 20)) %>%
  dplyr::select(uniprot, mode_plddt)
```

And the plot was created using the same design idea:

```{r plddt-mode-distro, fig.width=5, fig.align='center', eval=F}
ggplot(with_modes, aes(x = mode_plddt)) +
  geom_histogram(bins = 50, fill = "#1f77b4") +
  xlab("Modes of proteins (pLDDT)") +
  ylab("") +
  labs(title = "Distribution of mode pLDDTs") +
  theme(axis.line = element_line(colour = "black"),
        axis.text = element_text(size=14),
        axis.title = element_text(size=14),
        title = element_text(size=12),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        plot.margin = margin(t = 30),
        aspect.ratio = 0.75) +
  scale_x_continuous(breaks = c(0,20,40,60,80,100), limits = c(0,100)) +
  scale_y_continuous(expand = c(0,0,0.1,0)) +
  theme(plot.title = element_text(hjust = 0.5))
```

Finally, the IDPs are decided based on a cut-off of 60.

```{r, eval=F}
# first convert accessions:
ensembl_up_mapping %>%
  inner_join(with_modes, join_by(uniprotswissprot == uniprot)) %>%
  dplyr::select(ensembl_gene_id, mode_plddt) %>%
  write_tsv("IDP decisions/all_proteins_from_alphafold_with_mode.tsv")


# Filter by threshold 60
read_tsv("IDP decisions/all_proteins_from_alphafold_with_mode.tsv") %>%
  filter(mode_plddt < 60) %>%
  write_tsv("IDP decisions/idps_from_alphafold_mode.tsv")
```

