---
title: "DESeq Call with IDPs"
author: "Tolga Tabanli"
date: "2025-01-25"
output: pdf_document
knit: (function(input, encoding) {
  rmarkdown::render(input,
                    encoding = "UTF-8",
                    output_dir = "/home/tolga/Desktop/IDPs_under_Heatshock/Knittings")})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
knitr::opts_knit$set(root.dir = '/home/tolga/Desktop/IDPs_under_Heatshock/')
```

# DESeq Call with IDPs

This chapter summarizes how the experiment samples were mapped and
the DESeq was called accordingly. It also includes the creation of
volcano plots and the highlighting of the inferred intrinsically disordered proteins.

## Library Calls and Sample Preparations

```{r prep, message=FALSE}
library(DESeq2)
library(tidyverse)
library(ggpubr)
library(grid)

colData <- read_tsv("sample_mapping.tsv", col_names = T) %>%
  mutate(across(everything(), as.factor)) %>%
  column_to_rownames(var = "sample")

heatshock_counts <- read_tsv(here::here("complex_yeast_heatshock.tsv"), col_names = T) %>%
  column_to_rownames(var = "gene_id") %>%
  relocate(rownames(colData))

knockouts <- colData %>%
  select(knockout) %>%
  distinct() %>%
  pull() %>%
  relevel("Wildtype")

idps <- readRDS("IDP decisions/commons_modes.Rds")
deseq_results <- list()
```

```{r read-deseq , echo=FALSE}
deseq_results <- readRDS("deseq_results.Rds")
```

## Function Definitions for Multiple DESeq and Plotting

Firstly, the function that calls DESeq with given parameters is defined.
This function takes the genotype, temperature and time information
and carrious out DESeq with reference to temperature = 25 and time = 0 of
the respective genotype.

Importantly, this function uses a global assignment to store DESeq results
in a list outside of the function's scope. This way was preferred, because
the function's return value is used to create the grid plot.

```{r deseq-func}
multiple_deseq <- function(countData, colData, knock, temp, t) {
  subColData <- colData %>%
    filter(knockout == knock) %>%
    filter((temperature == 25 & time == 0) | temperature == temp)
  
  subsample <- subColData %>%
    filter(time %in% c(0, t))
  subcounts <- heatshock_counts %>% select(rownames(subsample))
  
  dds <- DESeqDataSetFromMatrix(countData = subcounts,
                                colData = subsample,
                                design = ~ time)
  deseq <- DESeq(dds, quiet = TRUE)
  res <- results(deseq)
  
  deseq_results[[paste0(knock,"_",temp,"_",t)]] <<- res
  
  return(plot_res(res, knock, temp, t))
}
```

Next function is used for creating an aesthetic grid plot that brings
together 16 different volcano plots.

```{r plot-func}
plot_res <- function(res, knock, temp, t) {
  res <- res %>%
    as.data.frame() %>%
    drop_na()
  significant_idps <- res$padj < 0.05 & rownames(res) %in% idps
  
  grob <- grobTree(textGrob(
    paste("Significant IDP ratio:\n",
          round(sum(significant_idps) / length(idps), digits = 2)),
    x=0.5,  y=0.6))
  
  p <- ggplot(data = res,
              aes(x = log2FoldChange, y = -log10(pvalue))) +
    scale_color_manual(values = c("#0400ff", "#747880"),
                       labels = c("Significant IDPs", "Others"),
                       breaks = c(TRUE, FALSE)) +
    geom_point(data = res %>% filter(!significant_idps), aes(colour = FALSE)) +
    geom_point(data = res %>% filter(significant_idps), aes(colour = TRUE)) +
    annotation_custom(grob) +
    labs(title = paste(knock, temp, "C, t =", t),
         color = "Significance")
  return(p)
}
```

## Multiple-DESeq Call

To analyze data, first the parameter space was created for the comparisons.

```{r parameters}
times <- c(10, 30)
temps <- c(37, 42)
knockouts <- colData %>%
  dplyr::select(knockout) %>%
  unique() %>%
  pull()

combs <- expand.grid(time = times, temperature = temps, knockout = knockouts)
plots <- list()
```

Next, we generate the DESeq results. While the actual results are being
stored in a global list, the individual plots are stored in "plots".
These plots were then plotted using the plot_res function for
labeling and highlighting IDP proteins.

```{r tutorial-shown, eval = FALSE}
for (row in 1:nrow(combs)) {
  plots[[row]] <- multiple_deseq(heatshock_counts,
                                 colData,
                                 combs[row,"knockout"],
                                 combs[row,"temperature"],
                                 combs[row, "time"])
}

for (ind in seq_along(deseq_results)) {
  plots[[ind]] <- plot_res(deseq_results[[ind]],
                           combs[ind, "knockout"],
                           combs[ind, "temperature"],
                           combs[ind, "time"])
}
```


Then, the plot was saved using ggarrange for fusion:

```{r fig-save, fig.height=25, fig.align='center', eval=F}
ggarrange(plotlist = plots, ncol = 2, nrow = 8, common.legend = T)
```

The same was repeated just to compare the effect of heat shock (42 °C)
on IDP expression, using wildtypes only.

```{r repeat-for-wildtype, eval=T, fig.height=3, fig.align='center'}
# For Heat shock DESeq with reference to wildtype (37) using only wildtypes
multiple_deseq_wt <- function(countData, colData, knock, temp, t) {
  subColData <- colData %>%
    filter(knockout == knock & time == t) %>%
    filter(temperature == 37 | temperature == temp) # REFERENCE AS 37
  
  subcounts <- heatshock_counts %>% select(rownames(subColData))
  
  dds <- DESeqDataSetFromMatrix(countData = subcounts,
                                colData = subsample,
                                design = ~ temperature)
  deseq <- DESeq(dds, quiet = TRUE)
  res <- results(deseq)
  
  deseq_results[[paste0(knock,"_",temp,"_",t)]] <<- res
  
  return(plot_res(res, knock, temp, t))
}

# reset results
deseq_results <- list()

times <- c(10, 30)
temps <- 42
knockouts <- "Wildtype"

combs <- expand.grid(time = times, temperature = temps, knockout = knockouts)
plots <- list()

for (row in 1:nrow(combs)) {
  plots[[row]] <- multiple_deseq(heatshock_counts,
                                 colData,
                                 as.character(combs[row,"knockout"]),
                                 combs[row,"temperature"],
                                 combs[row, "time"])
}

for (ind in seq_along(deseq_results)) {
  plots[[ind]] <- plot_res(deseq_results[[ind]],
                           combs[ind, "knockout"],
                           combs[ind, "temperature"],
                           combs[ind, "time"])
}

ggarrange(plotlist = plots, ncol = 2, nrow = 1, common.legend = T)
```

