---
title: "XXX"
author: "Tolga Tabanli"
date: "2025-01-25"
output: pdf_document
knit: (function(input, encoding) {
  rmarkdown::render(input,
                    encoding = "UTF-8",
                    output_dir = "/home/tolga/Desktop/IDPs_under_Heatshock/Knittings")})
---

```{r setup_enrichment, include=FALSE}
# DONE
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/home/tolga/Desktop/IDPs_under_Heatshock/')
```

Functional enrichment analysis was done with help of g:Profiler interface
gprofiler2. Dotplot was implemented using ggplot to facilitate query management,
as a direct conversion of gost() results to other packages such enrich was not
successful (multiquery requests would cause problems, another option would be
to vertically add results of single queries).

First, a function was defined the underlying dotplot data which are
the enrichment data for up and down regulation (reg) in each condition
(subdeseq), resulting in 16 sets.

```{r, message=F}
dotplot_updown <- function(sub_deseq, reg) {
  xdirection <- sub_deseq %>%
    filter(direction == reg & idp) %>%
    pull(rowname)
  multi_gp = gost(query = xdirection,
                  organism = "scerevisiae",
                  multi_query = FALSE, evcodes = TRUE)
  gp_mod = multi_gp$result[,c("query", "source", "term_id",
                              "term_name", "p_value", "query_size", 
                              "intersection_size", "term_size", 
                              "effective_domain_size", "intersection")]
  gp_mod$GeneRatio = gp_mod$intersection_size / gp_mod$query_size
  gp_mod$BgRatio = paste0(gp_mod$term_size, "/", gp_mod$effective_domain_size)
  names(gp_mod) = c("Cluster", "Category", "ID", "Description", "p.adjust", 
                    "query_size", "Count", "term_size", "effective_domain_size", 
                    "geneID", "GeneRatio", "BgRatio")
  gp_mod$geneID = gsub(",", "/", gp_mod$geneID)
  return(gp_mod)
}
```

Finally, the function was called for each condition in a for loop,
with plots saved and then arranged in a grid.
Importantly, I sort the gost() result according to increasing
term_size and decreasing Count, to get more specific terms that
are shared by as most of the genes as possible.
Here, only the GO-IDs are plotted due to size issues.
While concentrating on Wildtypes, the filter that was commented out
was also carried out.

```{r, message=F, fig.height=10}
dot_data <- list()
expr_names <- names(expr_direction)
for (cond in seq_along(expr_direction)*2) {
  label <- expr_names[[cond/2]]
  dot_data[[cond - 1]] <- dotplot_updown(expr_direction[[cond/2]], "Up") %>%
    mutate(Cluster = paste0(label, "_", "Up"))
  dot_data[[cond]] <- dotplot_updown(expr_direction[[cond/2]], "Down") %>%
    mutate(Cluster = paste0(label, "_", "Down"))
}

dot_data %>%
  map(filter, Category %in% c("GO:MF", "GO:BP", "GO:CC")) %>%
  map(arrange, term_size, desc(Count)) %>%
  map(slice_head, n = 10) %>%
  bind_rows() %>%
  #filter(str_starts(Cluster, "Wildtype")) %>%
  ggplot(aes(x = Cluster, y = ID, size = Count, color = p.adjust)) +
  geom_point() +
  scale_size_continuous(range = c(3, 12)) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=1)) +
  labs(x = "Cluster", y = "Enrichment Term", title = "Gene Enrichment Analysis") +
  facet_wrap(Category ~ ., scales = "free_y", ncol = 1)
```
