---
title: "Supporting Code for Analysis and Visualization"
author: "Tolga Tabanli"
date: "2025-02-15"
knit: (function(input, encoding) {
  rmarkdown::render(input,
                    encoding = "UTF-8",
                    output_dir = "/home/tolga/Desktop/IDPs_under_Heatshock/Knittings")})
output:
  pdf_document:
    toc: yes
---

```{r big-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/home/tolga/Desktop/IDPs_under_Heatshock/Report-Markdown')
```

# *Readme*

This document is an extensive summary of the main code snippets that
were used for the processing of RNAseq count data and the following
plot generation procedure.
Many code sections were omitted that were used for exploration,
trial-and-errors, accession mapping, and of course the Python code.
Python code can be accessed in
[GitHub Page](https://github.com/tolgatabanli/IDPs_under_Heatshock).
Plots used in the IDP Selection were also omitted for conciseness, and
can again be accessed in the GitHub page.

# IDP Decision

```{r child="aiupred_idps.Rmd"}
```


```{r child="alphafold_idps.Rmd"}
```

# DESeq

```{r child="deseq.Rmd"}
```

# Comparison of Differential Expressions

```{r child="comparison.Rmd"}
```

# Functional Enrichment Analysis

```{r child="enrichment.Rmd"}
```

# Session Info

```{r}
sessionInfo()
```

